Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-01 examples.tex
+

\subsubsection{Loading Page-Table Address Value}
In Figure \ref{fig:getnexttable}, we see how to obtain the physical address of the next level page-table. Traversal start from the root address of the page-table tree (the address of L4 table \textsf{pte\_t *pml4}) given as a parameter to \textsf{pte\_walkpgdir}). Then, we obtain the address of the entry (\textsf{pml4[PML4EX(va)]}) in the L4 table via the index in the virtual address (\textsf{va}).

Finally, we call \textsf{get\_next\_table} to get the virtual address to the next level page-table -- at this point the address of L3 table.
Within \textsf{get\_next\_table} first checks if the entry whose address is passed as a parameter ($\textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry}$). To be able to do so, it needs to load the entry value (a 64-bits word divided into bit-fields for corresponding fileds shown in Section {}) and check that \textit{presence} bit of the entry is set or not.

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta   }_{\rtv}$
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }\urcorner}$
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }  \urcorner }$
and    0x1,rdi
mov    rdi,rax
;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
lea    -0x10[rbp],rdi
;;pte_initialize(&next);
callq  70 <pte_initialize>
...
\end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}
\end{figure}

As we see in Figure \ref{fig:calltopteinitialize}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $\texttt{r8}$, but not the memory mapping to load the entry value referenced by the physical address \textsf{entry}. We need to construct this virtual mappings for \textsf{entry}.
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}
Now, to read the value referenced by physical address \textsf{entry} in Figure \ref{fig:calltopteinitialize}, first, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

 \subsubsection{Ensuring Presence of Entry}
 Accessing the presence bit and checking the value allow us to exploit the fact in the side condition ()such that we can own twe introduced for the  
In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module but the allocator's) to be backing the physical frame address of the entry.

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$

;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
lea    -0x10[rbp],rdi
;;pte_initialize(&next);
callq  70 <pte_initialize>
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
...

;;}
\end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}
\end{figure}

>>>>

\subsubsection{Loading Page-Table Address Value}
We call \textsf{get\_next\_table} to get the virtual address to the next level page-table. In Figure \ref{fig:calltopteinitialize}, we present the critical piece of \textsf{get\_next\_table}, and construct the specification of it 
Within \textsf{get\_next\_table} first checks if the entry whose address is passed as a parameter ($\textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry}$). To be able to do so, it needs to load the entry value (a 64-bits word divided into bit-fields for corresponding fileds shown in Section {}) and check that \textit{presence} bit of the entry is set or not.

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta   }_{\rtv}$
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }\urcorner}$
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }  \urcorner }$
and    0x1,rdi
mov    rdi,rax
;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
lea    -0x10[rbp],rdi
;;pte_initialize(&next);
callq  70 <pte_initialize>
...
;;}
;; uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn);
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs 0x0,rcx	                187: R_X86_64_64	KERN_BASE
add    rcx,rax
...
\end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}
\end{figure}

As we see in Figure \ref{fig:calltopteinitialize}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $\texttt{r8}$, but not the memory mapping to load the entry value referenced by the physical address \textsf{entry} (Line 7 in Figure \ref{fig:calltopteinitialize}). We need to construct this virtual mappings for \textsf{entry}.
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}
Before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

 \subsubsection{Ensuring Presence of Entry}
 Accessing the presence bit and checking the value allow (Lines 7-13 in Figure \ref{fig:calltopteinitialize}) us to exploit the fact in the side condition (*) such that we can own twe introduced for the  
In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module but the allocator's) to be backing the physical frame address of the entry.
